Delayed effects of [responses to?] climate lead to demographic divergence in a fragmented Amazonian landscape

### 2021-03-22 doc/paper.Rmd
work on discussion outline

@@

@@

Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, the dry season in the previous year).
Drought at lags 7--13 reduced flowering probability.
The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.

@@

    -   This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
    -   We found evidence for delayed effects of drought which are sometimes better predictors of vital rates than more recent conditions.
    -   Delayed effects of climate on plant vital rates may be the rule, rather than the exception.
    -   distributed lag non-linear models offer an unbiased way of identifying these delayed effects (but see @pierreCriticalWindowsMethod2020 for an alternative method). <!--# maybe the previous sentence goes in methods? -->

3.  What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat and life-history stage.
    [this lays out what these patterns are]

-   Growth v Survivorship v Flowering

@@

@@

-   Previous work has shown flowering probability to be an important vital rate.
    Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.

@@

@@

5.  Ecological Consequences of these patterns: an eye towards a drier future

>>>>

<!--# consider boxplots or median ± IQR for panel a) -->

@@

<!--# Is growth really higher in fragments?  Probably, since fig2 shows size becoming more similar over time.   -->

@@

Drought history had a significant effect on flowering probability in both habitat types (p \< 0.001).
Flowering probability was overall higher at all values of SPEI in continuous forest compared to 1-ha fragments (Figure \@ref(fig:flwr-spei)).
The effects of SPEI on flowering probability were also stronger and more complex in continuous forest compared to fragments (1-ha edf = 8.5, continuous forest edf = 10.5).
In continuous forest, flowering probability was increased by recent drought (at lag = 0 with SPEI \< -1) and drought at lags 15--20 (Jun-Nov, two dry seasons ago).
Drought at lags 7--13 reduced flowering probability.
The shape of the crossbasis function was similar in 1-ha fragments, but significantly muted (Figure \@ref(fig:flwr-spei)a).
That is, the variation in fitted values along the SPEI dimension was much lower at all values of the lag dimension in fragments.

@@

    This study adds to the growing body of work that finds delayed effects of climate drivers on vital rates without *a priori* assumptions about the influence of particular climate windows [@tenhumbergTimelaggedEffectsWeather2018; @tellerLinkingDemographyDrivers2016; @eversLaggedDormantSeason2021].
    We found evidence for delayed effects of drought which, in our models, are sometimes stronger predictors of vital rates than more recent conditions.
    Distributed lag non-linear models offer an unbiased way of identifying these delayed effects without overfitting (but see @pierreCriticalWindowsMethod2020 and @ogleQuantifyingEcologicalMemory2015a for alternative methods), yet few demographic studies have the data required for modeling lagged effects [@tellerLinkingDemographyDrivers2016; @croneHowPlantEcologists2011].

3.  What makes our results important is that we went the next step: we not only found that there were delayed effects, but that these effects different by habitat ~~and life-history stage~~ <!--# we didn't really do this.  This implies a size * SPEI interaction -->.
    [this lays out what these patterns are]

    -   Survival at mean SPEI was similar in fragments and continuous forest for all plant sizes.
        Plants were smaller in forest fragments, but grew slightly faster than plants in continuous forest, closing the size gap over the decade of data.

    -   However, the effect of varying SPEI on survival and growth was much greater in fragments.

    -   At all sizes, variation in SPEI had a greater impact on both growth and survival.
        Both unusually wet and unusually dry conditions resulted in reductions in survival in fragments, with lagged effects of up to around 15 months.
        Similarly for plant size, extremes in both directions in the census month caused reductions in growth in fragments.
        However, the lagged effects of SPEI on plant size depended on season with the general pattern of unusually dry wet seasons and unusually wet dry seasons being beneficial to plant growth.
        Together, this could possibly be explained by reduced microclimate buffering in forest fragments.
        More research... <!--# add more, possibly move this whole paragraph. -->

@@

    -   Our results confirm those of Bruna and Kress [-@brunaHabitatFragmentationDemographic2002] showing that probability of flowering increases with plant size, and is higher in continuous forests due to the greater number of large plants compared to fragments.
        <!--# combine with first bullet point. -->

@@

@@

    -   We also considered that some delayed effects could be a result of a cost of reproduction (triggered by an extreme SPEI) leading to reduced growth, survival, and/or flowering probability in the following season.
        However, when flowering in the previous season is included as a covariate, it predicts higher growth and flowering probability---likely a reflection of size being a good predictor of these vital rates (Supplemental\_\_\_).
        <!--# double-check against models and edit. -->

@@

5.  Ecological Consequences of these patterns: an eye towards a drier/wetter future
